# The Nelson–Aalen Estimator

The **Nelson–Aalen (NA) estimator** is a non-parametric estimator of the **cumulative hazard function** $H(t)$. Proposed by Wayne Nelson (1972) for reliability data and later formalised by Odd Aalen (1978) using counting-process theory, it provides an alternative to the Kaplan–Meier approach that is particularly useful in small samples and as a diagnostic tool.

```{r}
#| label: setup-na
#| echo: false
suppressPackageStartupMessages({
  library(survival); library(survminer); library(ggplot2)
  library(dplyr); library(knitr); library(kableExtra)
})
theme_set(theme_bw(base_size = 13))
data(lung)
lung$SurvObj <- Surv(lung$time, lung$status == 2)
```

## The Estimator

At each observed event time $t_j$ with $d_j$ events and $n_j$ subjects at risk:

$$\hat{H}(t) = \sum_{t_j \leq t} \frac{d_j}{n_j}$$

The corresponding survival estimate is then:

$$\tilde{S}(t) = \exp\{-\hat{H}(t)\}$$

This is sometimes called the **Fleming–Harrington estimator** in the survival package.

### Variance

$$\widehat{\text{Var}}[\hat{H}(t)] = \sum_{t_j \leq t} \frac{d_j}{n_j^2}$$

## Manual Calculation

```{r}
#| label: tbl-na-manual
#| tbl-cap: "Manual Nelson–Aalen calculation"

times  <- c(1, 2, 3, 3, 4, 5, 7, 8, 8, 10, 12, 14)
events <- c(1, 1, 1, 0, 1, 0, 1, 1, 0,  1,  0,  1)
uniq_et <- unique(sort(times[events == 1]))

na_tbl <- data.frame(t_j = uniq_et)
na_tbl$n_j  <- sapply(na_tbl$t_j, \(x) sum(times >= x))
na_tbl$d_j  <- sapply(na_tbl$t_j, \(x) sum(times == x & events == 1))
na_tbl$incr <- na_tbl$d_j / na_tbl$n_j
na_tbl$H_t  <- cumsum(na_tbl$incr)
na_tbl$S_tilde <- exp(-na_tbl$H_t)
na_tbl$var_H   <- cumsum(na_tbl$d_j / na_tbl$n_j^2)

na_tbl |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  kbl(col.names = c("t_j","n_j","d_j","d_j/n_j","Ĥ(t)","exp{-Ĥ(t)}","Var[Ĥ(t)]"),
      align = "c") |>
  kable_styling(bootstrap_options = c("striped","condensed"),
                full_width = FALSE, position = "center")
```

## Nelson–Aalen in R

```{r}
#| label: fig-na-lung
#| fig-cap: "Nelson–Aalen cumulative hazard estimate with confidence band"
#| fig-align: center
#| fig-width: 8
#| fig-height: 4

na_fit <- survfit(SurvObj ~ 1, data = lung, type = "fh")  # Fleming-Harrington

na_df <- data.frame(
  time   = na_fit$time,
  cumhaz = na_fit$cumhaz,
  lo     = na_fit$cumhaz - 1.96 * sqrt(na_fit$cumhaz / na_fit$n.risk),
  hi     = na_fit$cumhaz + 1.96 * sqrt(na_fit$cumhaz / na_fit$n.risk)
)

ggplot(na_df, aes(x = time, y = cumhaz)) +
  geom_step(colour = "#2C7BB6", linewidth = 1.2) +
  geom_ribbon(aes(ymin = pmax(0, lo), ymax = hi),
              fill = "#2C7BB6", alpha = 0.15) +
  labs(title = "Nelson–Aalen Cumulative Hazard Estimate",
       subtitle = "NCCTG Lung Cancer Data",
       x = "Time (days)", y = expression(hat(H)(t)))
```

## KM vs Nelson–Aalen: Which to Use?

```{r}
#| label: fig-km-vs-na
#| fig-cap: "KM and Nelson–Aalen survival estimates: differences emerge in small samples"
#| fig-align: center
#| fig-width: 8
#| fig-height: 4

km_fit  <- survfit(SurvObj ~ 1, data = lung)
na_surv <- exp(-na_fit$cumhaz)

df_comp <- bind_rows(
  data.frame(time = km_fit$time, surv = km_fit$surv, Method = "Kaplan–Meier"),
  data.frame(time = na_fit$time, surv = na_surv,      Method = "Nelson–Aalen")
)

ggplot(df_comp, aes(x = time, y = surv, colour = Method)) +
  geom_step(linewidth = 1.1) +
  scale_colour_manual(values = c("Kaplan–Meier" = "#2C7BB6",
                                 "Nelson–Aalen" = "#D7191C")) +
  labs(title = "KM vs Nelson–Aalen Survival Estimates",
       x = "Time (days)", y = "S(t)",
       colour = NULL) +
  theme(legend.position = "bottom")
```

::: {.callout-note title="When to Prefer Nelson–Aalen"}
The NA estimator is slightly **less biased** than KM in small samples (KM can be zero if the last observation is an event; NA never reaches zero). In practice, both give very similar results for large samples. The NA estimator is the natural output when cumulative hazards are of primary interest — e.g., when checking proportional hazards.
:::

::: {.callout-tip title="Quotes to Inspire"}
*"Without data, you're just another person with an opinion."*\
**— W. Edwards Deming**
:::
