# Concomitant Variables and Model Building

Selecting and interpreting covariate effects is as important as fitting the model. This chapter covers additive and multiplicative hazard models, age-of-onset distributions, and strategies for covariate selection in survival models.

```{r}
#| label: setup-cov
#| echo: false
suppressPackageStartupMessages({
  library(survival); library(survminer); library(ggplot2)
  library(dplyr); library(knitr); library(kableExtra); library(broom)
})
theme_set(theme_bw(base_size = 13))
data(lung)
```

## Multiplicative vs Additive Models

### Proportional Hazards (Multiplicative)

$$h(t;\mathbf{x}) = h_0(t) \exp(\boldsymbol{\beta}^T \mathbf{x})$$

Covariates act **multiplicatively** on the hazard. Log hazard ratio is linear in covariates.

### Aalen's Additive Hazard Model

$$h(t;\mathbf{x}) = h_0(t) + \boldsymbol{\beta}(t)^T \mathbf{x}$$

Covariates act **additively** on the hazard. The regression coefficients $\boldsymbol{\beta}(t)$ are time-varying, making this a very flexible model. Useful when the PH assumption fails.

## Selecting Concomitant Variables

### Step 1: Univariable Screening

Fit each covariate separately and flag those with $p < 0.25$ (a liberal threshold used to avoid excluding important predictors):

```{r}
#| label: tbl-univar
#| tbl-cap: "Univariable Cox models for covariate screening"

covs <- c("age","sex","ph.ecog","ph.karno","wt.loss")
uni_results <- lapply(covs, function(v) {
  fml <- as.formula(paste("Surv(time, status==2) ~", v))
  fit <- coxph(fml, data = lung)
  tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    select(term, estimate, conf.low, conf.high, p.value) |>
    mutate(across(where(is.numeric), \(x) round(x,3)))
})
bind_rows(uni_results) |>
  kbl(col.names = c("Covariate","HR","Lower 95%","Upper 95%","p-value"),
      align = "lcccc") |>
  kable_styling(bootstrap_options = c("striped","condensed"),
                full_width = FALSE, position = "center")
```

### Step 2: Multivariable Model with Stepwise Selection

```{r}
#| label: stepwise-cox

vars <- c("time","status","age","sex",
          "ph.ecog","ph.karno","wt.loss")

lung_cc <- lung[complete.cases(lung[, vars]), vars]

cox_full <- coxph(
  Surv(time, status == 2) ~ age + sex + ph.ecog +
    ph.karno + wt.loss,
  data = lung_cc
)

cox_step <- step(cox_full, direction = "both", trace = 0)

summary(cox_step)
```

### Step 3: Model Diagnostics

**Cox–Snell residuals** assess overall fit; **Martingale residuals** check functional form; **Deviance residuals** identify influential observations.

```{r}
#| label: fig-residuals
#| fig-cap: "Martingale residuals vs covariates to assess functional form"
#| fig-align: center
#| fig-width: 10
#| fig-height: 4

# Null model residuals for functional form assessment
cox_null <- coxph(Surv(time, status==2) ~ 1, data = lung)
mart_res  <- residuals(cox_null, type = "martingale")
lung_r    <- lung |> mutate(mart = mart_res)

p1 <- ggplot(lung_r, aes(x = age, y = mart)) +
  geom_point(alpha = 0.4, colour = "#2C7BB6") +
  geom_smooth(method = "loess", se = TRUE, colour = "#D7191C") +
  labs(title = "Age", x = "Age (years)", y = "Martingale Residual")

p2 <- ggplot(lung_r, aes(x = ph.karno, y = mart)) +
  geom_point(alpha = 0.4, colour = "#2C7BB6") +
  geom_smooth(method = "loess", se = TRUE, colour = "#D7191C") +
  labs(title = "Karnofsky Score", x = "Karnofsky", y = "Martingale Residual")

library(patchwork)
p1 + p2
```

## Logistic Regression for Fixed-Time Survival

When interest lies in survival to a **specific time point** $t_0$ (e.g., one year), logistic regression is appropriate:

$$\text{logit}[P(T \leq t_0 \mid \mathbf{x})] = \alpha + \boldsymbol{\beta}^T\mathbf{x}$$

```{r}
#| label: logistic-surv

lung_lr <- lung |>
  mutate(dead_1yr = as.integer(status == 2 & time <= 365))

logit_fit <- glm(
  dead_1yr ~ age + factor(sex) + ph.ecog,
  data = lung_lr,
  family = binomial
)

tidy(logit_fit, exponentiate = TRUE, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high, statistic, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kbl(
    col.names = c(
      "Covariate",
      "OR",
      "Lower 95%",
      "Upper 95%",
      "z",
      "p-value"
    ),
    caption = "Logistic regression: odds ratios for 1-year mortality",
    align = "lccccc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width = FALSE,
    position = "center"
  )
```

## Age-of-Onset Distributions

For diseases where the event is disease onset rather than death, we model:

$$S(t) = P(\text{onset age} > t)$$

A **parallel-system** model assumes $k$ susceptibility components must all "fail" for onset to occur:

$$S(t) = [S_1(t)]^k$$

If $S_1(t) = e^{-\lambda t}$ (exponential), then $S(t) = e^{-k\lambda t}$ — still exponential but with rate $k\lambda$. With Weibull components, the parallel model produces a Weibull mixture.

::: {.callout-tip title="Quotes to Inspire"}
*"It is easy to lie with statistics. It is hard to tell the truth without statistics."*\
**— Andrejs Dunkels**
:::
