# Interval Censoring

Interval censoring occurs when the event time is known only to lie within an interval $(L_i, R_i]$. This is extremely common in agricultural studies where inspections are periodic: a diseased plant found at week 5 means onset occurred *between* weeks 4 and 5.

```{r}
#| label: setup-ic
#| echo: false
suppressPackageStartupMessages({
  library(survival); library(ggplot2); library(dplyr)
  library(knitr); library(kableExtra); library(km.ci)
})
theme_set(theme_bw(base_size = 13))
```

## The Interval-Censored Likelihood

For observation $i$ with inspection interval $(L_i, R_i]$:

$$L_i(\theta) = S(L_i;\theta) - S(R_i;\theta) = P(L_i < T_i \leq R_i)$$

The full likelihood is $L(\theta) = \prod_{i=1}^n [S(L_i) - S(R_i)]$.

## The Turnbull Algorithm

The non-parametric MLE for interval-censored data is computed by the **Turnbull (1976)** self-consistency algorithm — an EM algorithm that identifies which intervals can support probability mass.

In R, `Surv(lower, upper, type = "interval2")` encodes interval-censored observations. The `survfit()` function uses the Turnbull algorithm automatically.

```{r}
#| label: fig-interval-km
#| fig-cap: "Non-parametric survival estimate from interval-censored data (Turnbull estimator)"
#| fig-align: center
#| fig-width: 8
#| fig-height: 5

set.seed(123)

n <- 80

icdat <- data.frame(
  lower = runif(n, 1, 8),
  upper = runif(n, 8, 15),
  treat = factor(rep(c("Control", "Treatment"), each = n/2))
)

icdat$upper <- pmax(icdat$upper, icdat$lower + 0.5)

ic_surv <- Surv(icdat$lower, icdat$upper, type = "interval2")

ic_fit <- survfit(ic_surv ~ treat, data = icdat)

plot(
  ic_fit,
  col = c("#2C7BB6", "#D7191C"),
  lwd = 2,
  xlab = "Time",
  ylab = "Survival Probability",
  main = "Turnbull Estimator for Interval-Censored Data"
)

legend(
  "topright",
  legend = levels(icdat$treat),
  col = c("#2C7BB6", "#D7191C"),
  lwd = 2,
  bty = "n"
)
```

::: {.callout-note title="Types of Interval Censoring"}
-   **Case I** (current-status data): each subject inspected once; we know only whether the event has occurred by inspection time
-   **Case II**: each subject inspected at multiple times; general $(L_i, R_i]$ intervals
-   Special cases: right censoring ($R_i = \infty$), exact observation ($L_i = R_i = t_i$), left censoring ($L_i = 0$)
:::
