# Competing Risks

In many survival studies, subjects can experience one of $k \geq 2$ distinct event types. Each type "competes" to occur first. A patient may die from cancer (cause 1), from cardiovascular disease (cause 2), or from other causes (cause 3). Occurrence of any one type precludes observation of the others.

```{r}
#| label: setup-cr
#| echo: false
suppressPackageStartupMessages({
  library(survival); library(ggplot2); library(dplyr)
  library(knitr); library(kableExtra)
})
theme_set(theme_bw(base_size = 13))
```

## Cause-Specific Hazard

The **cause-specific hazard** for cause $j$:

$$h_j(t) = \lim_{\Delta t \to 0} \frac{P(t \leq T < t+\Delta t,\; J=j \mid T \geq t)}{\Delta t}$$

The overall hazard is the sum: $h(t) = \sum_{j=1}^k h_j(t)$, and the overall survival:

$$S(t) = \exp\!\left\{-\sum_{j=1}^k H_j(t)\right\}$$

## Cumulative Incidence Function

The **cause-specific CIF** (also called the *sub-distribution function*):

$$F_j(t) = P(T \leq t,\, J = j) = \int_0^t S(u) h_j(u)\, du$$

Note: $\sum_j F_j(t) = 1 - S(t) \leq 1$.

::: {.callout-important title="Critical Warning: The 1-KM Mistake"}
A common error is to estimate $F_j(t)$ by applying $1 - \hat{S}_j(t)$ where $\hat{S}_j$ is the KM estimator treating all other event types as censored. This **overestimates** $F_j(t)$ because it ignores the fact that competing events remove subjects from the risk of cause $j$. The correct estimate is the cumulative incidence function (CIF), also called the **AalenŌĆōJohansen estimator**.
:::

## Indices for Probability of Death under Competing Risks

Two conceptually different quantities:

| Measure | Meaning |
|---------|---------|
| **Net probability** $Q_j(t)$ | Risk of dying from cause $j$ if cause $j$ were the *only* cause of death (hypothetical) |
| **Crude probability** $F_j(t)$ | Actual (observable) risk of dying from cause $j$ in the real world |

$$Q_j(t) = 1 - \exp\{-H_j(t)\}$$

## Computing CIFs in R

```{r}
#| label: fig-cif
#| fig-cap: "Cumulative incidence functions under competing risks (simulated data)"
#| fig-align: center
#| fig-width: 8
#| fig-height: 5

set.seed(123)
n <- 600
T1 <- rweibull(n, shape=1.5, scale=400)
T2 <- rweibull(n, shape=0.8, scale=600)
Tc <- runif(n, 200, 1000)

obs_t <- pmin(T1, T2, Tc)
cause <- ifelse(T1 == pmin(T1,T2,Tc), 1,
         ifelse(T2 == pmin(T1,T2,Tc), 2, 0))

cr_data  <- data.frame(time = obs_t, cause = cause)
cr_surv  <- Surv(cr_data$time, cr_data$cause, type = "mstate")
cif_fit  <- survfit(cr_surv ~ 1)

plot(cif_fit, col = c("#2C7BB6","#D7191C"), lwd = 2,
     xlab = "Time", ylab = "Cumulative Incidence",
     main = "Competing Risks: Cumulative Incidence Functions")
lines(survfit(Surv(obs_t, cause>0) ~ 1), col = "grey50", lwd=1.5, lty=2)
legend("topleft",
       legend = c("CIF Cause 1","CIF Cause 2","1 - S(t) (overall)"),
       col    = c("#2C7BB6","#D7191C","grey50"),
       lwd    = c(2,2,1.5), lty = c(1,1,2), bty = "n")
```

## Cause-Specific Cox vs FineŌĆōGray Models

Two approaches for regression with competing risks:

| Approach | Model | Quantity estimated |
|----------|-------|--------------------|
| **Cause-specific Cox** | Separate Cox for each cause; others treated as censored | Cause-specific hazard ratio |
| **FineŌĆōGray** | Subdistribution hazard model | Effect on CIF directly |

```{r}
#| label: cs-cox

cr_data$x1 <- rnorm(n); cr_data$x2 <- rbinom(n,1,0.5)
cr_data$ev1 <- as.integer(cr_data$cause==1)
cr_data$ev2 <- as.integer(cr_data$cause==2)

cs1 <- coxph(Surv(time, ev1) ~ x1 + x2, data = cr_data)
cs2 <- coxph(Surv(time, ev2) ~ x1 + x2, data = cr_data)

cat("=== Cause 1: Cause-specific Cox ===\n"); print(summary(cs1)$coef)
cat("\n=== Cause 2: Cause-specific Cox ===\n"); print(summary(cs2)$coef)
```

::: {.callout-tip title="Quotes to Inspire"}
*"Statistical thinking will one day be as necessary for efficient citizenship as the ability to read and write."*\
**ŌĆö H. G. Wells**
:::
