# Comparing Survival Experiences

A central question in many studies is whether two or more groups differ in their survival experience. This chapter presents the major non-parametric tests for comparing survival curves.

```{r}
#| label: setup-comp
#| echo: false
suppressPackageStartupMessages({
  library(survival); library(survminer); library(ggplot2)
  library(dplyr); library(knitr); library(kableExtra)
})
theme_set(theme_bw(base_size = 13))
data(lung)
```

## The Log-Rank Test

The **log-rank test** (Mantel–Cox, 1966) is the standard non-parametric test for comparing two or more survival curves. It is most powerful under the proportional hazards alternative.

### Two-Sample Log-Rank

At each observed event time $t_j$, construct a $2 \times 2$ table:

|  | Events | Non-events | Total |
|--|--------|-----------|-------|
| Group 1 | $O_{1j}$ | $n_{1j}-O_{1j}$ | $n_{1j}$ |
| Group 2 | $O_{2j}$ | $n_{2j}-O_{2j}$ | $n_{2j}$ |
| Total | $d_j$ | $n_j - d_j$ | $n_j$ |

**Expected events in group 1** at $t_j$: $E_{1j} = d_j \cdot n_{1j}/n_j$

**Test statistic:**

$$U = \sum_j (O_{1j} - E_{1j}), \qquad V = \sum_j \frac{n_{1j} n_{2j} d_j (n_j - d_j)}{n_j^2 (n_j - 1)}$$

$$\chi^2_{LR} = U^2 / V \sim \chi^2_1 \quad \text{under } H_0: S_1(t) = S_2(t)$$

### Mantel–Haenszel Form

The Mantel–Haenszel statistic is equivalent to the log-rank test:

$$\chi^2_{MH} = \frac{[\sum_j(O_{1j} - E_{1j})]^2}{\sum_j V_j}$$

## Weighted Tests

The general weighted log-rank statistic is:

$$\chi^2_w = \frac{[\sum_j w_j(O_{1j}-E_{1j})]^2}{\sum_j w_j^2 V_j}$$

| Test | Weights $w_j$ | Most powerful against |
|------|-------------|----------------------|
| Log-rank (Mantel-Cox) | $1$ | Proportional hazards |
| Wilcoxon (Breslow–Gehan) | $n_j$ | Early differences |
| Tarone–Ware | $\sqrt{n_j}$ | Intermediate |
| Fleming–Harrington($\rho$) | $[\hat{S}(t_j^-)]^\rho$ | Late differences ($\rho=1$) |

```{r}
#| label: logrank-tests

# Log-rank (rho = 0)
lr  <- survdiff(Surv(time, status==2) ~ sex, data = lung, rho = 0)
# Wilcoxon / Breslow (rho = 1)
wb  <- survdiff(Surv(time, status==2) ~ sex, data = lung, rho = 1)

cat("=== Log-Rank Test ===\n"); print(lr)
cat("\n=== Wilcoxon (Breslow) Test ===\n"); print(wb)
```

## k-Sample Test

For $k > 2$ groups, the generalised Mantel–Haenszel statistic:

$$\chi^2 = \mathbf{U}^T \hat{\boldsymbol{\Sigma}}^{-1} \mathbf{U} \sim \chi^2_{k-1}$$

```{r}
#| label: ksample-test

lr_ecog <- survdiff(Surv(time, status==2) ~ ph.ecog, data = lung)
print(lr_ecog)
```

```{r}
#| label: fig-km-ecog
#| fig-cap: "KM curves by ECOG performance status with log-rank p-value"
#| fig-align: center
#| fig-width: 9
#| fig-height: 7

lung2 <- lung |> filter(!is.na(ph.ecog))
km_ecog <- survfit(Surv(time, status==2) ~ ph.ecog, data = lung2)

ggsurvplot(km_ecog,
           data        = lung2,
           pval        = TRUE,
           pval.method = TRUE,
           conf.int    = FALSE,
           risk.table  = TRUE,
           legend.labs = paste("ECOG", sort(unique(lung2$ph.ecog))),
           palette     = c("#2C7BB6","#1A9641","#D7191C","#FF7F00"),
           xlab        = "Time (days)", ylab = "S(t)",
           title       = "Survival by ECOG Performance Status",
           ggtheme     = theme_bw(base_size = 13))
```

## Comparison of Two Life Tables

For two independent cohorts with complete mortality data, we test $H_0: q_{1x} = q_{2x}$ at each age $x$ using the normal approximation for proportions.

::: {.callout-note title="Choice of Test"}
Use the **log-rank test** when you expect proportional hazards (constant hazard ratio over time). Use **Wilcoxon** when group differences are largest early in follow-up. If curves cross (non-proportional hazards), no single weighted log-rank test is uniformly powerful — consider max-combo tests or the restricted mean survival time approach.
:::

::: {.callout-tip title="Quotes to Inspire"}
*"Errors using inadequate data are much less than those using no data at all."*\
**— Charles Babbage**
:::
