# Censoring and Truncation

Censoring is the defining feature of survival data. It arises whenever the exact event time is not observed for every subject — something that happens in virtually every real study. Handling censoring correctly is *the* central methodological challenge of survival analysis; ignoring it leads to systematically biased estimates.

```{r}
#| label: setup-cens
#| echo: false
suppressPackageStartupMessages({
  library(ggplot2); library(dplyr); library(survival)
  library(knitr); library(kableExtra)
})
theme_set(theme_bw(base_size = 13))
```

## Why Censoring Arises

Consider monitoring 100 insect colonies for the time until collapse. At the end of the study period:

- 68 colonies have collapsed (exact event times observed)
- 32 colonies are still active (we know only that $T_i > t_{\text{end}}$)

For the 32 surviving colonies, simply *ignoring* them understates colony longevity — we are throwing away real information. Treating their observation time as an event time (i.e., pretending they collapsed at $t_{\text{end}}$) overestimates mortality. Censoring methods allow us to use the partial information correctly.

## Types of Censoring

### Right Censoring (Most Common)

The event time $T_i$ exceeds the **observation time** $C_i$. We observe:

$$\tilde{T}_i = \min(T_i, C_i), \qquad \delta_i = I(T_i \leq C_i)$$

where $\delta_i = 1$ if the event is observed, $\delta_i = 0$ if censored.

**Sub-types:**

| Type | Mechanism | Example |
|------|-----------|---------|
| **Type I** | Study ends at fixed time $c$ | Field trial terminated at harvest |
| **Type II** | Study ends after $r$ events | Accelerated life test: stop at 10th failure |
| **Random** | Censoring time $C_i$ is random, independent of $T_i$ | Loss to follow-up; competing event |

### Left Censoring

The event occurred *before* observation began. We know only $T_i < t_0$.

> *Example*: A plant found to be already infested at first inspection — infestation time is left-censored at the inspection date.

### Interval Censoring

We know only that $L_i < T_i \leq R_i$ — the event occurred in a known interval.

> *Example*: Trees inspected weekly; disease detected at week 5 means onset was between weeks 4 and 5. (Covered in detail in Unit V.)

### Left Truncation (Delayed Entry)

A subject is only *eligible* for the study after surviving to some entry time $t_0$. Any failure before $t_0$ is unobservable — the subject would not have been recruited.

> *Example*: A retrospective study of crop lifespans: trees less than 5 years old are excluded. All included trees have *already survived* 5 years.

::: {.callout-important title="Informative vs Non-Informative Censoring"}
The standard analysis assumes **non-informative (independent) censoring**: the censoring time $C_i$ is independent of the failure time $T_i$ given any covariates. If patients who feel sicker are more likely to drop out, censoring is *informative* and standard methods are biased. This assumption must always be assessed in practice.
:::

## The Censored Data Likelihood

With right censoring, the correct likelihood contribution for each subject is:

$$L_i(\theta) = \begin{cases} f(t_i;\theta) & \text{if } \delta_i = 1 \text{ (event observed)} \\ S(t_i;\theta) & \text{if } \delta_i = 0 \text{ (censored)} \end{cases}$$

Combined: $L_i(\theta) = [f(t_i;\theta)]^{\delta_i} [S(t_i;\theta)]^{1-\delta_i} = [h(t_i;\theta)]^{\delta_i} S(t_i;\theta)$

The full likelihood is $L(\theta) = \prod_{i=1}^n L_i(\theta)$.

```{r}
#| label: fig-censoring
#| fig-cap: "Timeline diagram illustrating different censoring patterns for 10 subjects"
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

set.seed(42)
n <- 10
df_cens <- data.frame(
  id     = factor(n:1),
  start  = c(0,0,0,1,2,0,3,0,1,0),
  end    = c(3,6,7,8,5,4,10,9,6,2),
  status = c("Event","Censored","Event","Censored","Event",
             "Censored","Censored","Event","Event","Censored")
)

ggplot(df_cens, aes(y = id)) +
  geom_segment(aes(x = start, xend = end, yend = id),
               linewidth = 1.3, colour = "#2C7BB6") +
  geom_point(aes(x = end, colour = status, shape = status), size = 4) +
  geom_vline(xintercept = 10, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("Event" = "#D7191C","Censored" = "#636363")) +
  scale_shape_manual(values  = c("Event" = 4, "Censored" = 1)) +
  annotate("text", x = 10.15, y = 0.5, label = "Study end", hjust = 0, size = 3.5) +
  labs(title = "Survival Study Timeline: Events and Censored Observations",
       x = "Time", y = "Subject ID",
       colour = "Outcome", shape = "Outcome") +
  theme(legend.position = "bottom")
```

## Creating Survival Objects in R

The `survival` package represents censored data with a `Surv` object. For right-censored data:

```{r}
#| label: surv-object
library(survival)

# Simulate 15 observations: failure times and censoring indicators
set.seed(101)
time   <- round(rexp(15, rate = 0.1), 1)   # true failure times
censor <- runif(15, 3, 12)                  # censoring times
obs_t  <- pmin(time, censor)               # observed time
delta  <- as.integer(time <= censor)        # event indicator

# Create Surv object
surv_obj <- Surv(time = obs_t, event = delta)
print(surv_obj)  # '+' marks censored observations
```

::: {.callout-tip title="Reading Surv Output"}
In the printed `Surv` object, a `+` after a time value means the observation is right-censored — the event had not yet occurred at that time. Exact event times appear without any suffix.
:::

## Representing the Three Common Censoring Types in R

```{r}
#| label: surv-types

# Right censoring (standard)
s_right <- Surv(time = obs_t, event = delta, type = "right")

# Left censoring: event known to have occurred before observation time
# event = 1 when time is LEFT-censored (event already happened)
left_t     <- c(2, 5, 3, 8, 4)
left_event <- c(1, 0, 1, 0, 1)   # 1 = left-censored, 0 = observed
s_left <- Surv(time = left_t, event = left_event, type = "left")

# Interval censoring: lower and upper bounds
lower_t <- c(1, 3, 2, 5, 4)
upper_t <- c(3, 6, 4, 8, 7)
s_interval <- Surv(time = lower_t, time2 = upper_t, type = "interval2")

cat("Right-censored:\n"); print(s_right[1:6])
cat("\nLeft-censored:\n");  print(s_left)
cat("\nInterval-censored:\n"); print(s_interval)
```

::: {.callout-tip title="Quotes to Inspire"}
*"The most important questions of life are indeed, for the most part, really only problems of probability."*\
**— Pierre-Simon Laplace**
:::
